---
title: "The mysterious corrleation of BF and *p* under H0"
output: 
  rmdformats::html_clean:
    highlight: kate
---

```{r knitr_init, echo=FALSE, cache=FALSE, message=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               comment=NA,
               tidy=FALSE,
               message=TRUE,
               warning=FALSE)
opts_knit$set(width=75)
```

*NB: This is my first ever RMarkdown script! I expect it to be all kinds of faulty, please let me know about improvements.*

First we install a few packages
```{r, eval = FALSE}
install.packages("BayesFactor")
install.packages("compute.es")
install.packages("ggplot2")
install.packages("ggExtra")
```

# Background

It all started with a tweet by Eiko Fried, who was puzzled about a high correlation between Bayes factors (BF) and *p*-values although BFs are said to be a "continuous measure of evidence" and *p*-values are not. This is because BFs compare evidence for H1 with evidence for H0. The smaller a BF10, the more (relative) evidence for H0, the larger it is, the more (relative) evidence for H1. *P*-values are different: When H0 is false, they are right-skewed (more small *p*-values, fewer large ones), but when H0 is true, they are uniformly distributed (all values between 0 and 1 are equally likely, the distribution is flat).

Felix Schönbrodt pointed this out and explained that that probably meant BF and *p* would be highly correlated when H0 is false but not when H0 is true. I thought this idea was amazingly enlightening and tweeted a screenshot of the conversation: <https://twitter.com/annemscheel/status/867419377582432258>

# The Rebuttal

Unfortunately the idea turned out to be false. Richard Morey jumped in and made his point in an ingenious tweet that included his point *and* code to prove it: <https://twitter.com/richarddmorey/status/867439728257363971>

Let's go through Richard's code bit by bit. 
The first chunk simulates 10,000 *t*-values from the central distribution (this means that H0 is true and the values we get will be centred on 0) with 50 degrees of freedom and accompanying *p*-values. (Note that we didn't set a seed, so you will get slightly different values every time you run this script.)
```{r TP}
t=rt(1e4,50) 
p=2*pt(-abs(t),50) 
```

The second bit uses the BayesFactor package and generates BFs for the *t*-values we just simulated. It will take a moment to compile.

```{r BF, message=FALSE}
library("BayesFactor")
b=exp(sapply(t,function(t) 
  BayesFactor::ttest.tstat(t,51)$bf)) 
```

Now we have 10,000 *p*-values and their respective BFs, drawn from a population in which the null is true. The *p*-distribution should be flat and the BF-distribution shouldn't, so surely they are not correlated, right...? Let's have a look. This bit gives a scatterplot of the log of BF and *p*:

```{r BFplot}
plot(p,b,log="xy") 
```
Wow! Well, that was unexpected. 

Here is the correlation of log(p) and log(b):
```{r BFcor}
cor(log(p),log(b))
```

# Understanding, pt. 1

I first found this puzzling, but then I realised: Just like BFs, *p*-values track evidence under the null. You will get large *p*-values for small effects and small *p*-values for large effects. The BF works the other way round and a bit differently, but the main principle is the same (smaller effects provide more evidence for H0). As you saw above, the code even generates *p*-values and BFs from the same *t*-values! So clearly they must have something in common.

Ok. I got that. But now came befuddlement number 2: *P*-values "track" the effect size under H0. But effect size is not uniformly distributed under the null, it is normally distributed: Many small values and few extreme ones.  
**Why the hell is the *p*-distribution flat under the null?**

# Understanding, pt. 2

I spent a good while plotting the data in all kinds of ways until I understood. The whole process felt very embarrassing, because this is such a basic and trivial aspect of *p*-values. But hey, that insight doesn't magically put knowledge into your head, so I guess it's better to understand this now than never, even if I wouldn't want my first-year stats lecturer to see this.

First I calculated effect sizes (*d*) from the simulated t-distribution, because I feel more familiar with *d*. Then I created a plot that showed both the distribution of effect size and the corresponding *p*-distribution.

We now calculate *d* and then additionally the absolute value of *d* to get a second plot with only positive values for the effect size distribution. This code is probably far from perfect - it's pretty slow and I'm sure there's a better way to solve this.


```{r unicornmat}
library("compute.es")
d <- sapply(t,function(t)
  compute.es::tes(t,26,26, verbose=FALSE)$d)
absd <- abs(d)
unicornmat <- cbind(d, absd, p)
unicornmat <- as.data.frame(unicornmat)
```

# Scatterplot of *d* and *p* with marginal histograms
Now we're ready to plot! We first create a scatterplot of *d* and *p* using ggplot2 and then add marginal histograms using ggExtra.

```{r unicornplot, message=FALSE}
library("ggplot2")
library("ggExtra")

unicornplot <- ggplot(unicornmat, aes(unicornmat$d, unicornmat$p)) + 
  geom_point(shape=1) +
  theme(text = element_text(size=15)) +
  labs(title="Effect size (d) versus p-values") +
  labs(x="effect size (d)", y="p-value")

unicornplot <- ggExtra::ggMarginal(unicornplot, type = "histogram")
unicornplot
```

You see that the *p*-value "tracks" the effect size, which is normally distributed, but the overall *p*-distribution is still flat! That's  because the same difference in *d* will give you a larger difference in *p* for small effect sizes and a smaller difference in *p* for large effect sizes. In other words: There are way fewer d-values in the tails of the distribution, but they are "concentrated" in very small bins on the *p*-distribution. I hope the plot gets the point across.

And I'm embarrassed about this whole ordeal because *p*-values were of course constructed to behave in this exact way! 

# Scatterplot of absolute *d* and *p* with marginal histograms
Last plot of the day: We make the same plot as above, just with absolute values for *d*. In other words, all negative *d*s were added onto the positive side of the distribution.

```{r absoluteunicornplot, message=FALSE}
absoluteunicornplot <- ggplot(unicornmat, aes(unicornmat$absd, unicornmat$p)) +
  geom_point(shape=1) + 
  theme(text = element_text(size=15)) +
  labs(title="Absolute effect size (d) versus p-values") +
  labs(x="absolute effect size (d)", y="p-value")

absoluteunicornplot <- ggExtra::ggMarginal(absoluteunicornplot, type = "histogram")
absoluteunicornplot
```
